---
title: "Extrapolating BCI species distributions using Thomas point pattern"
author: "Petr Keil"
date: "December 9, 2015"
output: 
  html_document: 
    highlight: tango
    theme: cerulean
---

# Loading the necessary R libraries

```{r, message=FALSE}
  library(raster)
  library(sp)
  library(maptools)
  library(rasterVis)
  library(RColorBrewer)
  library(spatstat)
  library(plyr)
```

# Loading the data

I am using the first BCI census.

```{r, fig.width=12, fig.height=8}
  bci <- read.table("../data/PlotDataReport12-09-2015_1179045564.txt", header=T, sep="\t")
  # removing the NAs
  bci <- bci[is.na(bci$PX)==FALSE,]
  bci <- bci[, c("Latin","PX","PY")]
```


# The BCI plot

Here I plot the BCI plot in a hypothetical 2x2 km area, which will be used
for extrapolation of species distributions.

```{r, fig.width=10, fig.height=10}
    my.green <- col2rgb(brewer.pal(9, "Greens")[8])
    my.green <- rgb(my.green[1]/256, my.green[2]/256, my.green[3]/256, alpha=0.1)
    
    par(oma=c(0,0,0,0))
    plot(c(0,2000),c(0, 2000), type="n", axes=FALSE, xlab="x [m]", ylab="y [m]")
    axis(1); axis(2)
    points(bci$PX, bci$PY, cex=0.1, col=my.green) 
    rect(0,0,2000,2000)
```

# Function fitting the Thomas process model

The function returns parameters that can be used in the `rThomas` function.

```{r}
  Thomas <- function(x, y) # x and y coordinates of the points
  {
    if(length(x)==1) # in case there is just a single point
    {
      return(c(kappa=1/(1000*500), sigma=1, mu=1))  
    }
    
    else # fit the Thomas model and return the parameters
    {
      W <- as.owin(list(xrange=c(0,1000), yrange=c(0,500)))
      xy <- as.ppp(cbind(x, y), W)
      fit.xy <- clusterfit(xy, clusters="Thomas")
      return(fit.xy$modelpar)
    }
  }
```

Test the function

```{r, fig.width=10, fig.height=10}
  test.sp <- bci[bci$Latin=="Acalypha diversifolia",]
  #test.sp <- bci[bci$Latin=="Ormosia amazonica",]
  #test.sp <- bci[bci$Latin=="Bactris major",]
  #test.sp <- bci[bci$Latin=="Ficus popenoei",]
  #test.sp <- bci[bci$Latin=="Hybanthus prunifolius",]
  
  test.fit <- Thomas(test.sp$PX, test.sp$PY)
  
  W2000 <- as.owin(list(xrange=c(0,2000), yrange=c(0,2000)))
  
  test.pred <- rThomas(test.fit['kappa'], 
                       test.fit['sigma'], 
                       test.fit['mu'], 
                       win=W2000)

  plot(test.pred, pch=19, cex=0.3, cols="black", main="Test species")
  rect(0,0,1000,500, col="white")
  points(test.sp$PX, test.sp$PY, col="red", pch=19, cex=0.3)
```

Get species-abundance distribution of the original BCI plot:

```{r}
  abund <- ddply(.data=bci,
                 .variables="Latin",
                 .fun=summarize,
                  Abundance=length(Latin))
```

Fit the Thomas model to all species in the BCI plot, and export the results
as a data frame:

```{r}
  Thomas.res <- list()
  for(i in 1:nrow(abund))
  {
    spec <- as.character(abund[i, 'Latin'])
    bci.sub <- bci[bci$Latin==spec,]
    Thomas.res[[i]] <- c(Thomas(bci.sub$PX, bci.sub$PY))
  }  
  Thomas.res <- ldply(Thomas.res)
  Thomas.res <- data.frame(abund, Thomas.res)
```

Plot histograms of the fitted Thomas model parameters:

```{r, fig.width=10, fig.height=10}
  par(mfrow=c(2,2), mai=c(0.8, 0.8, 0.1, 0.1))
  hist(log10(Thomas.res$Abundance), col="grey", main="", 
       xlab="log10 Abundance", breaks=30)
  hist(log10(Thomas.res$kappa), col="grey", main="", 
       xlab="log10 kappa", breaks=30)
  hist(log10(Thomas.res$sigma), col="grey", main="", 
       xlab="log10 sigma", breaks=30)
  hist(log10(Thomas.res$mu), col="grey", main="", 
       xlab="log10 mu", breaks=30)
  pairs(log10(Thomas.res[,2:5]))
```

# Calculate SAD of the extrapolated 2x2 km community:

```{r}

```

# Exact versions of R and its packages used 

```{r}
  sessionInfo()
```



