---
title: "Extrapolating BCI species distributions using Thomas point pattern"
author: "Petr Keil"
output: 
  html_document: 
    highlight: tango
    theme: cerulean
---

# Loading the necessary R libraries

```{r, message=FALSE}
  library(raster)
  library(sp)
  library(maptools)
  library(rasterVis)
  library(RColorBrewer)
  library(spatstat)
  library(plyr)
  library(vegan)
```

# Loading the data

I am using the first BCI census.

```{r, fig.width=12, fig.height=8}
  bci <- read.table("../data/PlotDataReport12-09-2015_1179045564.txt", header=T, sep="\t")
  # removing the NAs
  bci <- bci[is.na(bci$PX)==FALSE,]
  bci <- bci[, c("Latin","PX","PY")]
```


# The BCI plot

Here I plot the BCI plot in a hypothetical 2x2 km area, which will be used
for extrapolation of species distributions.

```{r, fig.width=10, fig.height=10}
    my.green <- col2rgb(brewer.pal(9, "Greens")[8])
    my.green <- rgb(my.green[1]/256, my.green[2]/256, my.green[3]/256, alpha=0.1)
    
    par(oma=c(0,0,0,0))
    plot(c(0,2000),c(0, 2000), type="n", axes=FALSE, xlab="x [m]", ylab="y [m]")
    axis(1); axis(2)
    points(bci$PX, bci$PY, cex=0.1, col=my.green) 
    rect(0,0,2000,2000)
```

# Function fitting the Thomas process model

The function returns parameters that can be used in the `rThomas` function.

```{r}
  Thomas <- function(x, y) # x and y coordinates of the points
  {
    if(length(x)==1) # in case there is just a single point
    {
      return(c(kappa=1/(1000*500), sigma=1, mu=1))  
    }
    
    else # fit the Thomas model and return the parameters
    {
      W <- as.owin(list(xrange=c(0,1000), yrange=c(0,500)))
      xy <- as.ppp(cbind(x, y), W)
      fit.xy <- clusterfit(xy, clusters="Thomas")
      return(fit.xy$modelpar)
    }
  }
```

Function that fits, extrapolates and plots the Thomas model for a specific species:

```{r}
  Thomas.plot <- function(spec.name, bci)
  {
    sp <- bci[bci$Latin==spec.name,]
    sp.fit <- Thomas(sp$PX, sp$PY)
    W2000 <- as.owin(list(xrange=c(0,2000), yrange=c(0,2000)))
      
    sp.pred <- rThomas(sp.fit['kappa'], 
                       sp.fit['sigma'], 
                       sp.fit['mu'], 
                       win=W2000)
    
    plot(sp.pred, pch=19, cex=0.3, cols="black", main=spec.name)
    rect(0,0,1000,500, col="white")
    points(sp$PX, sp$PY, col="red", pch=19, cex=0.3)
  }
```

Example species

```{r, fig.width=10, fig.height=15}
par(mfrow=c(3,2), mai=c(0.1,0.1, 0.5, 0.5))
  Thomas.plot("Hybanthus prunifolius", bci)
  Thomas.plot("Acalypha diversifolia", bci)
  Thomas.plot("Bactris major", bci)
  Thomas.plot("Triplaris cumingiana", bci)
  Thomas.plot("Ficus popenoei", bci)
  Thomas.plot("Ormosia amazonica", bci)
  
```

Get species-abundance distribution of the original BCI plot:

```{r}
  abund <- ddply(.data=bci,
                 .variables="Latin",
                 .fun=summarize,
                  Abundance=length(Latin))
```

Fit the Thomas model to all species in the BCI plot, and export the results
as a data frame:

```{r, warning=FALSE}
  Thomas.res <- list()
  for(i in 1:nrow(abund))
  {
    spec <- as.character(abund[i, 'Latin'])
    bci.sub <- bci[bci$Latin==spec,]
    Thomas.res[[i]] <- c(Thomas(bci.sub$PX, bci.sub$PY))
  }  
  Thomas.res <- ldply(Thomas.res)
  Thomas.res <- data.frame(abund, Thomas.res)
```

Calculate SAD of the extrapolated 2x2 km community:

```{r}
  # define the 2x2 km window 
  W2000 <- as.owin(list(xrange=c(0,2000), yrange=c(0,2000)))
  abund.ext <- numeric() # empty container for the results

  for(i in 1:nrow(abund))
  {
    spec <- as.character(abund[i, 'Latin'])
    spec.2km <- rThomas(kappa=Thomas.res[i, 'kappa'], 
                        scale=Thomas.res[i, 'sigma'], 
                        mu=Thomas.res[i, 'mu'], 
                        win=W2000)
    abund.ext[i] <- spec.2km$n
  }  

  Thomas.res <- data.frame(Thomas.res, Abundance.ext=abund.ext)
```

Export the results to `.csv` file:

```{r}
  write.csv(Thomas.res, file="BCI_Thomas.csv", row.names = FALSE)
```

Plot abundance distributions:

```{r, fig.width=10, fig.height=10}
  par(mfrow=c(2,2), mai=c(0.8, 0.8, 0.1, 0.1))
  hist(log10(Thomas.res$Abundance), col="grey", main="", 
       xlab="log10 Abundance (500 x 1000)", breaks=30)
  hist(log10(Thomas.res$Abundance.ext), col="grey", main="", 
       xlab="log10 Abundance (2000 x 2000)", breaks=30)
  plot(sort(log10(Thomas.res$Abundance), decreasing=TRUE), 
       ylab="Log10 Abundance (500 x 1000)")
    plot(sort(log10(Thomas.res$Abundance.ext), decreasing=TRUE), 
       ylab="Log10 Abundance (2000 x 2000)")
```

Plot histograms of the fitted Thomas model parameters:

```{r, fig.width=10, fig.height=10}
  par(mfrow=c(2,2), mai=c(0.8, 0.8, 0.1, 0.1))
  hist(log10(Thomas.res$kappa), col="grey", main="", 
       xlab="log10 kappa", breaks=30)
  hist(log10(Thomas.res$sigma), col="grey", main="", 
       xlab="log10 sigma", breaks=30)
  hist(log10(Thomas.res$mu), col="grey", main="", 
       xlab="log10 mu", breaks=30)
```

```{r}
  pairs(log10(Thomas.res[,2:6]))
```



# Exact versions of R and its packages used 

```{r}
  sessionInfo()
```



